rbcL ASV clustering

### 2019-10-31 code/rbcl_asv_clustering.Rmd
new version of asv clustering, copied over from old asv_viz

---
title: "ASV visualization"
---

```{r libs, echo=F}
library(tidyverse)
library(here)
library(DECIPHER)
library(ggforce)
library(Rtsne)
library(ape)
library(ggtree)
library(treeio)
library(dbscan)
library(ggthemes)
library(ggpubr)
# library(plotly)
# library(crosstalk)
# library(DT)
```

```{r set}
theme_set(theme_few())
set.seed(123)
```

```{r read}
rbcl_asv_id_key <- read_csv(here("output","keys","rbcl_asv_key.csv"))
rbcl_rel_abund <- read_csv(here("data","clean","rbcl_rel_abund_clean.csv"))
idtax <- read_csv(here("data","clean","rbcl_taxonomy_clean.csv"))
```

Align rbcL ASVs
```{r align}
# view alignments, make distance matrix
asv.dss <- rbcl_asv_id_key$asv %>%
  DNAStringSet()
names(asv.dss) <- rbcl_asv_id_key$asv_id
asv.aligns <- asv.dss %>% 
  DECIPHER::AlignSeqs(verbose=F)

asv.aligns %>% DECIPHER::BrowseSeqs(highlight = T) # 600 has a ~70 bp insert, 607 ~45 bp deletion, 508 ~110 bp deletion
length(asv.aligns)
```

Make distance matrix
```{r dists}
dists <-  asv.aligns %>% 
  DECIPHER::DistanceMatrix(type="matrix", correction = "JC")
```

t-SNE with distance matrix
```{r tsne}
perplexity <- 30

@@

ggplot(tsne, aes(x=V1,y=V2))+
  geom_point()
```

Add taxonomy assignments, see if t-SNE agrees with the reference-based clustering of IDTaxa
```{r genus_tsne}
tsne <- tsne %>% 
  add_column(asv_id = colnames(dists)) %>% # label each point by it's ASV ID
  left_join(idtax, by="asv_id")

ggplot(tsne, aes(x=V1,y=V2))+
  geom_point(aes(color = genus))
```
Seperates out the Sanguia, Chlainomonas. Lumps all Trebouxiophyceae together. Lots of clustering within Chloromonas, some big clusters with only one point assigned to Chloromonas.

check at species level
```{r sp_tsne}
ggplot(tsne, aes(x=V1,y=V2))+
  geom_point(aes(color = species))
```

Make it look nice, show both genus and species level assignments
```{r tsne-plot1}

@@

tsne_plot <- ggplot(tsne, aes(x=V1, y=V2)) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne, is.na(display_taxa)), 
             color = "grey",alpha = 0.8, size = 3) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne, !is.na(display_taxa)), 
             aes(color = display_taxa %>% fct_relevel(taxa.order), alpha = 0.8), size = 3)  
tsne_plot
```

cluster t-SNE into OTUs
```{r otu-cluster}
eps <- 4

@@

@@

ggplot(tsne, aes(x=V1, y=V2)) +
  geom_point(aes(color = otu))
```

manually add *Chloromonas krienitzii* cluster, since this is highly abundant, and split *Chlainomonas* into two, based on the observation that the two clades of *Chlainomonas* seperate out by elevation
```{r edit-otus}
tsne <- tsne %>%
  mutate(otu = case_when((V2 < (-5) & V1 < -17) ~ "9", # chlain "high"
                         (V2 < 14  & V2 > 8 & V1 > (-20) & V1 < (-12)) ~ "10",# krienitzii
                         TRUE ~ otu))

@@

tsne_plot <- tsne_plot +
  ggforce::geom_mark_ellipse(data = tsne, aes(fill=otu_name, label=otu_name))

# set

# extract the legend
tsne_plot %>% 
  get_legend() %>% 
  as_ggplot()
ggsave(here("figs","asv-cluster","asv-legend.pdf"), height = 7, width = 3)

# plot and save the main plot
tsne_plot +
  theme(legend.position = "none")
ggsave(here("figs","asv-cluster","asv-cluster.pdf"), height = 5, width = 5)
```

@@

Run IQtree
```{r}
rbcl.tree <- read.tree(here("output","tree","rbcl","asv613.fasta.treefile"))

>>>>

---
title: "rbcL ASV clustering"
output: html_notebook
---

```{r libs, include=F}
library(tidyverse)
library(here)
library(DECIPHER)
library(ggforce)
library(Rtsne)
library(ape)
library(ggtree)
library(treeio)
library(dbscan)

library(ggpubr)
library(ggthemes)
library(ggdark)
```

```{r read, include=F}
asv_key <- read_csv(here("data/asv_seq_data/rbcl_asv_sequence_key.csv"))
rel_abund <- read_csv(here("data/03_filtered/rbcl/filtered_cleaned_rbcl_rel_abund.csv"))
taxonomy <- read_csv(here("data/02_tidied/tidied.rbcl_taxonomy.csv"))
```

```{r asv_distance_matrix, include=F}

aligns <- asv_key %>%
  filter(asv_id %in% rel_abund$asv_id) %>% # only want asvs used in the main analysis
  deframe() %>% 
  DNAStringSet() %>% 
  DECIPHER::AlignSeqs(verbose=F)

# check alignments
aligns %>% DECIPHER::BrowseSeqs(highlight = T) # ASV 639 has a ~70 bp insert, 646 ~45 bp deletion, 535 ~110 bp deletion
length(aligns)

dists <-  aligns %>% 
  DECIPHER::DistanceMatrix(type="matrix", correction = "JC")
```

```{r tsne_1, echo=F}
perplexity <- 30

set.seed(123)

@@

ggplot(tsne, aes(x=V1,y=V2))+
  geom_point() +
  labs(title="rbcL ASV clustering with t-SNE")
```

```{r genus_tsne, echo=F}
tsne <- tsne %>% 
  add_column(asv_id = colnames(dists)) %>% # label each point by it's ASV ID
  left_join(taxonomy, by="asv_id")

ggplot(tsne, aes(x=V1,y=V2))+
  geom_point(aes(color = genus)) +
  labs(title = "Points labeled by genus")
```

```{r sp_tsne, echo=F}
ggplot(tsne, aes(x=V1,y=V2))+
  geom_point(aes(color = species)) +
  labs(title="Pts labeled by species")
```

Make it look nice, show both genus and species level assignments
```{r tsne-plot1, echo=F}

@@

ggplot(tsne, aes(x=V1, y=V2)) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne, is.na(display_taxa)), 
             color = "grey",alpha = 0.8, size = 3) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne, !is.na(display_taxa)), 
             aes(color = display_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
  theme_few()
```

```{r otu-cluster, echo=F}
eps <- 4

@@

set.seed(123)

@@

ggplot(tsne, aes(x=V1, y=V2)) +
  geom_point(aes(color = otu)) +
  labs(title="dbscan clustering")
```

Manually edit the clustering:
  - *Chloromonas krienitzii* gets its own cluster, since this is highly abundant
  - split *Chlainomonas* into two, based on the observation in IQTree the two clades of *Chlainomonas* seperate out by elevation

```{r edit-otus, echo=F}
tsne <- tsne %>%
  mutate(otu = case_when((V2 < (-5) & V1 < -17) ~ "9", # chlain "high"
                         (V2 < 14  & V2 > 8 & V1 > (-20) & V1 < (-12)) ~ "10",# krienitzii
                         TRUE ~ otu))

@@

ggplot(tsne, aes(x=V1, y=V2)) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne, is.na(display_taxa)), 
             color = "grey",alpha = 0.8, size = 3) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne, !is.na(display_taxa)), 
             aes(color = display_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
  theme_few() +
  ggforce::geom_mark_ellipse(data = tsne, aes(fill=otu_name, label=otu_name), show.legend = FALSE)

```

```{r include_rel_abund, echo=F, fig.width=8, fig.height=6}
total_rel_abund <- rel_abund %>% 
  group_by(asv_id) %>% 
  summarise(global_rel_abund = sum(rel_abund))

tsne_2 <- tsne %>% 
  left_join(total_rel_abund, by="asv_id")

ggplot(tsne_2, aes(x=V1, y=V2)) +
  ggforce::geom_mark_ellipse(data = tsne, aes(fill=otu_name, label=otu_name), show.legend = FALSE) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne_2, is.na(display_taxa)), 
             aes(size = global_rel_abund),
             color = "grey",alpha = 0.7) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne_2, !is.na(display_taxa)), 
             aes(color = display_taxa %>% fct_relevel(taxa.order),
                 size = global_rel_abund), 
             alpha = 0.7) +
  labs(title = "rbcL ASV t-SNE clustering", color="ASV taxonomic assignment") +
  theme_few() +
  scale_size(range = c(1, 6), guide="none")
```

```{r tsne-output, echo=F}

tsne_plot +
  theme(legend.position = "none")

# extract the legend
tsne_plot %>% 
  get_legend() %>% 
  as_ggplot

ggsave(here("figs","asv-cluster","asv-legend-dark.pdf"), height = 7, width = 3)
ggsave(here("figs","asv-cluster","asv-cluster-dark.pdf"), height = 5, width = 5)
```

```{r asv_mds, echo=F}

```

@@

Run IQtree
```{r tree, fig.width=8, fig.height=20}
rbcl.tree <- read.tree(here("output","tree","rbcl","asv613.fasta.treefile"))
